### 2025-03-04 README.qmd
inital commit of rrtools compendium

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# WAlingex

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/nickgestrich/WAlingex/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2025-03-04 analysis/paper/paper.qmd
add language and location info

---
title: "Loan word maps"
author:
  - Henning Schreiber:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - uhh
  - Nikolas Gestrich:
      institute: fi
      orcid: 0000-0003-1689-0558

institute:
  - uhh:
      name: Universität Hamburg
      address: 23 Science Street, Eureka, Mississippi, USA
  - fi: Frobenius Institute
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Creating language maps

Read location data from glottolog and Sahelia databases

```{r}
#| label: get-data
#| eval: false
# Read data
langloc <- readr::read_csv(here::here('analysis/data/raw_data/language_location.csv')) |>

# drop languages with unknown location
  tidyr::drop_na(longitude) |>

  # turn latitude and longitude columns into geometry
  sf::st_as_sf(coords = c("longitude","latitude"))

# set crs
sf::st_crs(langloc) <- 4326
```

```{r}
# read classified loan word data

# Define the file path
file_path <- here::here('analysis/data/raw_data/cleaned exports.xlsx')

# Get the names of all the sheets in the Excel file
sheet_names <- readxl::excel_sheets(file_path)

# Read each sheet and store it as a separate data frame
for (sheet in sheet_names) {
  assign(sheet, readxl::read_excel(file_path, sheet = sheet, col_names = TRUE)) 
}

```

add the location data

```{r}
# add location to the gold sheet

# merge by glottolog code 
gold <- gold |> 
  #join by glottolog code
  dplyr::left_join(langloc, by = c("lang" = "GlottCode"))

# split off where glottolog code matching was successful
gold_a <-  gold |> dplyr::filter(!is.na(LanName))

gold_a <- gold |> filter(!is.na(LanName))

#where the glottolog code is not available, merge by Sahelia name
gold <- gold |> 
  dplyr::mutate(id2 = ifelse(is.na(LanName), id2, NA))

# Perform the first join on the primary id1
merged_data <- data1 %>%
  left_join(data2, by = c("id1" = "id"))

# For the rows where the primary ID did not match, perform a second join on id2
final_data <- merged_data %>%
  mutate(id2 = ifelse(is.na(x), id2, NA)) %>%   # Set id2 as NA if the primary join succeeded
  left_join(data2, by = c("id2" = "id")) %>%
  select(-id2)  # Drop the temporary id2 column
```

```{r}
library(dplyr)
library(stringr)
library(sf)
library(readr)
library(leaflet)
library(purrr)
library(glue)
library(htmltools)

# 3. Load classified language data and plot

# Load data
gold <- read_csv("gold.csv") %>%

  # Modify the "lang" column to keep only the first expression and remove the number behind it
  mutate(lang = str_trim(str_extract(lang, "^[^ ]+"))) %>%

  # Join the langloc dataset with the gold dataset
  inner_join(langloc, by = c("lang" = "lang_code")) %>%

  # Exclude rows where the "form" variable contains NA values
  filter(!is.na(form))

# 4. Plot on an interactive leaflet map
# Define a color palette for the "form" column
color_pal <- colorFactor(palette = "Set3", domain = gold$form)

# Add a new column with the corresponding colors
gold <- gold %>%
  mutate(color = color_pal(form))

# Create the leaflet map
leaflet(gold) %>%
  addProviderTiles(providers$Esri.WorldPhysical) %>%
  addLabelOnlyMarkers(
    lng = ~st_coordinates(geometry)[,1],
    lat = ~st_coordinates(geometry)[,2],
    label = purrr::map(
      glue::glue("<span style='color:{gold$color}; font-weight: bold; font-size: 12px;'>{as.character(gold$form)}</span>"),
      htmltools::HTML
    ),
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = 'top',
      textOnly = TRUE
    )
  )

## NEXT: run this for a list of files and output a html file leaflet map for each.

```

Merge using two columns

```{r}
# Perform the first join on the primary id1
merged_data <- data1 %>%
  left_join(data2, by = c("id1" = "id"))

# For the rows where the primary ID did not match, perform a second join on id2
final_data <- merged_data %>%
  mutate(id2 = ifelse(is.na(x), id2, NA)) %>%   # Set id2 as NA if the primary join succeeded
  left_join(data2, by = c("id2" = "id")) %>%
  select(-id2)  # Drop the temporary id2 column
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2025-03-12 analysis/paper/paper.qmd
added data and map production code

@@
 # Read data
-langloc <- readr::read_csv(here::here('analysis/data/raw_data/language_location.csv')) |>
+langloc <- readr::read_csv(here::here('analysis/data/raw_data/glottologAfricanLanguages.csv')) |>
+
+  #retain only relevant variables
+  dplyr::select(glott_code, latitude, longitude, LanName) |>
 
@@
 # read classified loan word data
+# Create a vector of file paths
+objs <- here::here("analysis", "data", "raw_data", c("gold.csv", 
+                                                     "fonio.csv",
+                                                     "millet.csv",
+                                                     "salt.csv",
+                                                     "iron.csv"))
 
-# Define the file path
-file_path <- here::here('analysis/data/raw_data/cleaned exports.xlsx')
+# read the csv files into a list of data frames
+data_list <- purrr::map(objs, readr::read_csv)
 
-# Get the names of all the sheets in the Excel file
-sheet_names <- readxl::excel_sheets(file_path)
+# Add the location data to the word lists
 
-# Read each sheet and store it as a separate data frame
-for (sheet in sheet_names) {
-  assign(sheet, readxl::read_excel(file_path, sheet = sheet, col_names = TRUE)) 
+# Function to join location data to each data frame
+join_geometry <- function(df) {
+  df |> 
+    dplyr::left_join(langloc, by = "glott_code") |> 
+    sf::st_as_sf()  # Convert the result to an sf object if needed
 }
 
+# Apply the join function to each data frame in the list
+data_list <- purrr::map(data_list, join_geometry)
 ```
 
-add the location data
+Function to make interactive leaflet maps
+```{r}
+create_leaflet_map <- function(data_sf, form_column = "form", background = "colour") {
 
+  # Ensure the input data is an sf object and contains the 'form' column
+  if (!inherits(data_sf, "sf")) {
+    stop("The input data must be an sf object.")
+  }
+  if (!form_column %in% colnames(data_sf)) {
+    stop(paste("The column", form_column, "is not found in the data."))
+  }
+
+  # Filter out rows with NA values in the "form" column
+  data_sf <- data_sf |>
+    dplyr::filter(!is.na(!!rlang::sym(form_column)))
+
+  # Set the tile provider based on the 'background' argument
+  tile_provider <- switch(
+    background,
+    "colour" = leaflet::providers$Esri.WorldPhysical,  # Default color tile
+    "bw" = leaflet::providers$Esri.WorldGrayCanvas,    # Black-and-white tile
+    stop("Invalid value for 'background'. Choose 'colour' or 'bw'.")
+  )
+
+  # Define a color palette for the "form" column (using Set3 palette)
+  color_pal <- leaflet::colorFactor(palette = "Set3", domain = data_sf[[form_column]])
+
+  # Add a new column with the corresponding colors based on the 'form' column
+  data_sf <- data_sf |>
+    dplyr::mutate(color = color_pal(!!rlang::sym(form_column)))  # Dynamically reference the column
+
+  # Adjust overlapping points by slightly jittering their coordinates
+  # Initially apply fixed jitter for the map creation
+  jitter_scale <- 0.1000  # Set the base jitter scale factor
+
+  data_sf <- data_sf |>
+    dplyr::mutate(
+      # Apply a small random jitter to coordinates for initial display
+      jittered_lon = sf::st_coordinates(geometry)[, 1] + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale),
+      jittered_lat = sf::st_coordinates(geometry)[, 2] + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale)
+    )
+
+  # Create the leaflet map
+  leaflet_map <- leaflet::leaflet(data_sf) |>
+    leaflet::addProviderTiles(tile_provider) |>
+    leaflet::addLabelOnlyMarkers(
+      lng = ~jittered_lon,  # Use jittered longitude
+      lat = ~jittered_lat,  # Use jittered latitude
+      label = purrr::map(
+        glue::glue("<span style='color:{data_sf$color}; font-weight: bold; font-size: 12px;'>{as.character(data_sf[[form_column]])}</span>"),
+        htmltools::HTML
+      ),
+      labelOptions = leaflet::labelOptions(
+        noHide = TRUE,
+        direction = 'top',
+        textOnly = TRUE
+      )
+    )
+
+  # Add JavaScript to dynamically adjust jitter based on zoom level
+  leaflet_map <- leaflet_map |>
+    htmlwidgets::onRender(
+      "
+      function(el, map) {
+        // Function to apply jitter based on zoom level
+        function applyJitter() {
+          var zoomLevel = map.getZoom();
+          var jitterScale = 0.1 * Math.pow(1.7, zoomLevel); // Adjust jitter with zoom level
+
+          // Adjust the jitter of each marker
+          map.eachLayer(function(layer) {
+            if (layer instanceof L.Marker) {
+              var latLng = layer.getLatLng();
+              var newLat = latLng.lat + (Math.random() - 0.5) * jitterScale;
+              var newLng = latLng.lng + (Math.random() - 0.5) * jitterScale;
+              layer.setLatLng([newLat, newLng]);
+            }
+          });
+        }
+
+        // Apply jitter on map zoom change
+        map.on('zoomend', function() {
+          applyJitter();
+        });
+
+        // Apply jitter initially
+        applyJitter();
+      }
+      "
+    )
+
+  # Return the map object
+  return(leaflet_map)
+}
+```
+
+Function to save a version of every map in the list of data
 ```{r}
-# add location to the gold sheet
+# Function to create and save leaflet maps for all data frames in data_list
+save_maps <- function(data_list, form_column = "form", background = "colour") {
 
-# merge by glottolog code 
-gold <- gold |> 
-  #join by glottolog code
-  dplyr::left_join(langloc, by = c("lang" = "GlottCode"))
+  # Use purrr::map2 to iterate over the list of data frames and generate maps
+  purrr::map2(data_list, seq_along(data_list), ~ {
+    # Create the map using the create_leaflet_map function
+    map <- create_leaflet_map(.x, form_column, background)
 
-# split off where glottolog code matching was successful
-gold_a <-  gold |> dplyr::filter(!is.na(LanName))
+    # Generate the filename based on the index or data frame name
+    filename <- paste0("map_", .y, ".html")
 
-gold_a <- gold |> filter(!is.na(LanName))
+    # Save the map as an HTML file in the 'analysis/figures' folder
+    htmlwidgets::saveWidget(map, here::here("analysis/figures", filename))
+  })
+}
+```
 
-#where the glottolog code is not available, merge by Sahelia name
-gold <- gold |> 
-  dplyr::mutate(id2 = ifelse(is.na(LanName), id2, NA))
-
-# Perform the first join on the primary id1
-merged_data <- data1 %>%
-  left_join(data2, by = c("id1" = "id"))
-
-# For the rows where the primary ID did not match, perform a second join on id2
-final_data <- merged_data %>%
-  mutate(id2 = ifelse(is.na(x), id2, NA)) %>%   # Set id2 as NA if the primary join succeeded
-  left_join(data2, by = c("id2" = "id")) %>%
-  select(-id2)  # Drop the temporary id2 column
+Save an html version of all objects
+```{r}
+save_maps(data_list, background = "bw")
 ```

### 2025-03-12 analysis/paper/paper.qmd
remove old code

@@
 ```{r}
-library(dplyr)
-library(stringr)
-library(sf)
-library(readr)
-library(leaflet)
-library(purrr)
-library(glue)
-library(htmltools)
-
-# 3. Load classified language data and plot
-
-# Load data
-gold <- read_csv("gold.csv") %>%
-
-  # Modify the "lang" column to keep only the first expression and remove the number behind it
-  mutate(lang = str_trim(str_extract(lang, "^[^ ]+"))) %>%
-
-  # Join the langloc dataset with the gold dataset
-  inner_join(langloc, by = c("lang" = "lang_code")) %>%
-
-  # Exclude rows where the "form" variable contains NA values
-  filter(!is.na(form))
-
-# 4. Plot on an interactive leaflet map
-# Define a color palette for the "form" column
-color_pal <- colorFactor(palette = "Set3", domain = gold$form)
-
-# Add a new column with the corresponding colors
-gold <- gold %>%
-  mutate(color = color_pal(form))
-
-# Create the leaflet map
-leaflet(gold) %>%
-  addProviderTiles(providers$Esri.WorldPhysical) %>%
-  addLabelOnlyMarkers(
-    lng = ~st_coordinates(geometry)[,1],
-    lat = ~st_coordinates(geometry)[,2],
-    label = purrr::map(
-      glue::glue("<span style='color:{gold$color}; font-weight: bold; font-size: 12px;'>{as.character(gold$form)}</span>"),
-      htmltools::HTML
-    ),
-    labelOptions = labelOptions(
-      noHide = TRUE,
-      direction = 'top',
-      textOnly = TRUE
-    )
-  )
-
-## NEXT: run this for a list of files and output a html file leaflet map for each.
-
-```
-
-Merge using two columns
-
-```{r}
-# Perform the first join on the primary id1
-merged_data <- data1 %>%
-  left_join(data2, by = c("id1" = "id"))
-
-# For the rows where the primary ID did not match, perform a second join on id2
-final_data <- merged_data %>%
-  mutate(id2 = ifelse(is.na(x), id2, NA)) %>%   # Set id2 as NA if the primary join succeeded
-  left_join(data2, by = c("id2" = "id")) %>%
-  select(-id2)  # Drop the temporary id2 column
-```
-
-```{r}
 #| label: fig-demo-plot

### 2025-03-17 analysis/paper/paper.qmd
updates to leaflet code

@@
 
+The data contains certain duplicate entries. These need to be removed to avoid overly full plots
+```{r}
+# Function to remove duplicates
+remove_duplicates <- function(df_list) {
+  df_list <- lapply(df_list, function(df) {
+    df <- df |> 
+      dplyr::distinct(glott_code, form, .keep_all = TRUE)
+    return(df)
+  })
+  return(df_list)
+}
+
+# Apply function
+data_list <- remove_duplicates(data_list)
+```
+
 Function to make interactive leaflet maps
+
 ```{r}
-create_leaflet_map <- function(data_sf, form_column = "form", background = "colour") {
+create_leaflet_map <- function(data_sf, form_column = "form", style_column = "style", background = "colour") {
 
-  # Ensure the input data is an sf object and contains the 'form' column
+  # Ensure the input data is an sf object and contains the required columns
   if (!inherits(data_sf, "sf")) {
@@
   }
-  if (!form_column %in% colnames(data_sf)) {
-    stop(paste("The column", form_column, "is not found in the data."))
+  if (!style_column %in% colnames(data_sf)) {
+    stop(paste("The column", style_column, "is not found in the data."))
+  }
+  if (!"glott_code" %in% colnames(data_sf)) {
+    stop("The column 'glott_code' is not found in the data.")
+  }
+  if (!"LanName" %in% colnames(data_sf)) {
+    stop("The column 'LanName' is not found in the data.")
   }
 
-  # Filter out rows with NA values in the "form" column
+  # Filter out rows with NA values in the "style" column
   data_sf <- data_sf |>
-    dplyr::filter(!is.na(!!rlang::sym(form_column)))
+    dplyr::filter(!is.na(!!rlang::sym(style_column)))
 
@@
 
-  # Define a color palette for the "form" column (using Set3 palette)
-  color_pal <- leaflet::colorFactor(palette = "Set3", domain = data_sf[[form_column]])
+  # Define a color palette for the "style" column (using Dark2 palette from RColorBrewer)
+  color_pal <- leaflet::colorFactor(palette = RColorBrewer::brewer.pal(8, "Dark2"), domain = data_sf[[style_column]])
 
-  # Add a new column with the corresponding colors based on the 'form' column
+  # Add a new column with the corresponding colors based on the 'style' column
   data_sf <- data_sf |>
-    dplyr::mutate(color = color_pal(!!rlang::sym(form_column)))  # Dynamically reference the column
+    dplyr::mutate(color = color_pal(!!rlang::sym(style_column)))  # Dynamically reference the column
 
@@
   # Initially apply fixed jitter for the map creation
-  jitter_scale <- 0.1000  # Set the base jitter scale factor
+  jitter_scale <- 0.5000  # Set the base jitter scale factor
 
@@
     dplyr::mutate(
-      # Apply a small random jitter to coordinates for initial display
-      jittered_lon = sf::st_coordinates(geometry)[, 1] + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale),
-      jittered_lat = sf::st_coordinates(geometry)[, 2] + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale)
+      original_lon = sf::st_coordinates(geometry)[, 1],  # Store original longitude
+      original_lat = sf::st_coordinates(geometry)[, 2],  # Store original latitude
+      jittered_lon = original_lon + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale),
+      jittered_lat = original_lat + stats::rnorm(dplyr::n(), mean = 0, sd = jitter_scale)
     )
@@
       )
+    ) |>
+    leaflet::addMarkers(
+      lng = ~jittered_lon,
+      lat = ~jittered_lat,
+      popup = ~paste("<strong>Glottocode:</strong> ", glott_code, "<br><strong>Language Name:</strong> ", LanName),
+      options = leaflet::markerOptions(opacity = 0)
     )
@@
           var zoomLevel = map.getZoom();
-          var jitterScale = 0.1 * Math.pow(1.7, zoomLevel); // Adjust jitter with zoom level
+          var jitterScale = 0.5 * Math.pow(20, -zoomLevel); // Increase jitter at lower zoom levels
 
@@
             if (layer instanceof L.Marker) {
-              var latLng = layer.getLatLng();
-              var newLat = latLng.lat + (Math.random() - 0.5) * jitterScale;
-              var newLng = latLng.lng + (Math.random() - 0.5) * jitterScale;
+              var originalLat = layer.options.originalLat;  // Access the original latitude
+              var originalLng = layer.options.originalLng;  // Access the original longitude
+              var newLat = originalLat + (Math.random() - 0.5) * jitterScale;
+              var newLng = originalLng + (Math.random() - 0.5) * jitterScale;
               layer.setLatLng([newLat, newLng]);
@@
       }
-      "
+      ",
+      list(
+        originalLat = ~original_lat,
+        originalLng = ~original_lon
+      )
     )

### 2025-03-17 analysis/paper/paper.qmd
update paper code

@@
       orcid: 0000-0003-1689-0557
-      institute:
-        - uhh
+      institute: uhh
+
   - Nikolas Gestrich:
@@
 #| label: get-data
-#| eval: false
+#| eval: true
 # Read data
@@
 
-```{r}
+```{r read data}
 # read classified loan word data
@@
 The data contains certain duplicate entries. These need to be removed to avoid overly full plots
-```{r}
+```{r clean duplicates}
 # Function to remove duplicates
@@
 
-```{r}
+```{r map function}
 create_leaflet_map <- function(data_sf, form_column = "form", style_column = "style", background = "colour") {
@@
 Function to save a version of every map in the list of data
-```{r}
+```{r save function}
 # Function to create and save leaflet maps for all data frames in data_list
@@
 ```{r}
+#| label: save-maps
+#| eval: false
 save_maps(data_list, background = "bw")

### 2025-03-19 analysis/paper/paper.qmd
update paper code
add reference

@@
 ---
-title: "Loan word maps"
+title: "Make maps of loan words"
 author:
@@
 
-# Creating language maps
-
-Read location data from glottolog and Sahelia databases
+In order to have location data, we read in the African languages, their glottoglog code, and their location from the glottolog database (@hammarstromGlottolog2024).
 
@@
 ```
+
+Next, we read in the data and join the entries to the glottolog entries.
+Modify the list of .csv files to suit your interests.
+If you want to upload new files, place these in the directory /analysis/data/raw_data.Make sure any new data has a glottolog language code in a varaible named "glott_code"
 
@@
 
-The data contains certain duplicate entries. These need to be removed to avoid overly full plots
+The data may contain certain duplicate entries. These need to be removed to avoid overly full plots. This is done by running the code below.
 ```{r clean duplicates}
@@
 #| eval: false
-save_maps(data_list, background = "bw")
+#save_maps(data_list, background = "bw")
+
+#create_leaflet_map(data_list[[2]])
 ```
-
-```{r}
-#| label: fig-demo-plot
-#| fig-cap: "A plot of random numbers"
-plot(rnorm(10))
-```
-
-@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
-
-```{r}
-#| label: demo-inline-code
-#| echo: false
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
-
-# Discussion
-
-# Conclusion
-
-# Acknowledgements
 

### 2025-03-20 analysis/paper/paper.qmd
paper and data update

--- 2025-03-17 analysis/paper/paper.qmd
@@
   These are the highlights. 
+prefer-html: true
 ---
@@
 
+```{r}
+#| label: fonio
+#| fig-cap: "fonio word forms"
+create_leaflet_map(data_list[[2]])
+```
+
 Save an html version of all objects
@@
 save_maps(data_list, background = "bw")
+
+create_leaflet_map(data_list[[2]])
 ```

### 2025-03-25 analysis/paper/paper.qmd
updates 24-3

--- 2025-03-19 analysis/paper/paper.qmd
@@
 data_list <- purrr::map(data_list, join_geometry)
+
+# Assign names
+names(data_list) <- c("gold", "fonio", "millet", "salt", "iron")
 ```
@@
 The data may contain certain duplicate entries. These need to be removed to avoid overly full plots. This is done by running the code below.
+
 ```{r clean duplicates}
@@
 ```
+
+Classifications go into more detail than we need for plotting. We will only retain a rough classification here:
+
+```{r}
+data_list <- data_list |> 
+  purrr::map(~ dplyr::mutate(.x, style = stringr::str_extract(style, "^[A-Za-z]")))
+```
+
+## Make interactive maps
 
@@
 Function to save a version of every map in the list of data
+
 ```{r save function}
@@
 
+```{r}
+#| label: fonio
+#| fig-cap: "fonio word forms"
+ create_leaflet_map(data_list[[1]], background = "bw")
+```
+
 Save an html version of all objects
+
 ```{r}
@@
 #| eval: false
-#save_maps(data_list, background = "bw")
+save_maps(data_list)
+```
 
-#create_leaflet_map(data_list[[2]])
+## Make static maps (for printing etc)
+
+Function to make static plots
+
+```{r}
+create_static_map <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {
+
+  # Ensure the input data is an sf object and contains the required columns
+  if (!inherits(data_sf, "sf")) {
+    stop("The input data must be an sf object.")
+  }
+  if (!style_column %in% colnames(data_sf)) {
+    stop(paste("The column", style_column, "is not found in the data."))
+  }
+  if (!"glott_code" %in% colnames(data_sf)) {
+    stop("The column 'glott_code' is not found in the data.")
+  }
+  if (!"LanName" %in% colnames(data_sf)) {
+    stop("The column 'LanName' is not found in the data.")
+  }
+
+  # Filter out rows with NA values in the "style" column
+  data_sf <- data_sf |> dplyr::filter(!is.na(!!rlang::sym(style_column)))
+
+  # Define the color palette for the "style" column, ensuring darker colors
+  num_categories <- length(unique(data_sf[[style_column]]))
+  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)
+
+  # Create the base map with background
+  base_map <- ggplot2::ggplot() +
+    ggspatial::annotation_map_tile(tiles = "http://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/${z}/${y}/${x}") +
+    ggplot2::geom_sf(data = data_sf, ggplot2::aes(color = !!rlang::sym(style_column)), show.legend = FALSE) +
+    ggplot2::scale_color_manual(values = dark_palette) +
+    ggplot2::theme_minimal() +
+    ggplot2::theme(legend.position = "bottom") +
+    ggplot2::labs(color = style_column)
+
+  # Add the plot title if provided
+  if (!is.null(plot_title)) {
+    base_map <- base_map + ggplot2::ggtitle(plot_title)
+  }
+
+  # Crop the extent of the map to the data layers' extent
+  base_map <- base_map + ggplot2::coord_sf(xlim = sf::st_bbox(data_sf)[c("xmin", "xmax")],
+                                           ylim = sf::st_bbox(data_sf)[c("ymin", "ymax")])
+
+  # Plot the data as strings contained in the form_column variable
+  map_plot <- base_map +
+    ggplot2::geom_text(data = data_sf, ggplot2::aes(x = sf::st_coordinates(data_sf)[,1],
+                                                    y = sf::st_coordinates(data_sf)[,2],
+                                                    label = !!rlang::sym(form_column),
+                                                    color = !!rlang::sym(style_column)),
+                       position = ggplot2::position_jitter(width = 0.1, height = 0.1), size = 3)
+
+  # Return the ggplot object
+  return(map_plot)
+}
+```
+
+Make static plots
+
+```{r}
+create_static_map(data_list[[2]], plot_title = "fonio")
+```
+
+```{r}
+create_static_map_tmap <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {
+
+  # Ensure the input data is an sf object and contains the required columns
+  if (!inherits(data_sf, "sf")) {
+    stop("The input data must be an sf object.")
+  }
+  if (!style_column %in% colnames(data_sf)) {
+    stop(paste("The column", style_column, "is not found in the data."))
+  }
+  if (!"glott_code" %in% colnames(data_sf)) {
+    stop("The column 'glott_code' is not found in the data.")
+  }
+  if (!"LanName" %in% colnames(data_sf)) {
+    stop("The column 'LanName' is not found in the data.")
+  }
+
+  # Filter out rows with NA values in the "style" column
+  data_sf <- data_sf |> dplyr::filter(!is.na(!!rlang::sym(style_column)))
+
+  # Define the color palette for the "style" column, ensuring darker colors
+  num_categories <- length(unique(data_sf[[style_column]]))
+  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)
+
+  # Get the background map using maptiles package
+  background_map <- maptiles::get_tiles(data_sf, provider = "Esri.WorldGrayCanvas", crop = TRUE)
+
+  # Increase the boundaries of the plot
+  bbox <- sf::st_bbox(data_sf)
+  x_range <- bbox["xmax"] - bbox["xmin"]
+  y_range <- bbox["ymax"] - bbox["ymin"]
+  bbox["xmin"] <- bbox["xmin"] - 50 * x_range
+  bbox["xmax"] <- bbox["xmax"] + 50 * x_range
+  bbox["ymin"] <- bbox["ymin"] - 50 * y_range
+  bbox["ymax"] <- bbox["ymax"] + 50 * y_range
+
+  # Create the tmap object
+  tmap::tmap_mode("plot")
+  map_plot <- tmap::tm_shape(background_map) +
+    tmap::tm_rgb() +
+    tmap::tm_shape(data_sf) +
+    tmap::tm_text(text = form_column, col = style_column, palette = dark_palette) +
+    tmap::tm_layout(title = plot_title, legend.show = FALSE, bbox = bbox)
+
+  # Return the tmap object
+  return(map_plot)
+}
+```
+
+```{r}
+create_static_map_tmap(data_list[[1]])
+```
+
+## Make a network
+
+Our network will be made using languages as nodes, which come with locations. We will then use the data in the list to create edges between the nodes.
+
+```{r}
+# Load necessary libraries
+library(dplyr)
+library(tidyr)
+
+# Function to create edge list for a dataframe
+create_edge_list <- function(df, df_name) {
+  # Clean and transform the data
+  df <- df |> 
+    select(glott_code, style) |> 
+    as_tibble()
+
+  # Create edge list based on first letter of style
+  edge_list_first_letter <- df |>
+    mutate(first_letter = substr(style, 1, 1)) |>
+    inner_join(., ., by = "first_letter") |>
+    filter(glott_code.x != glott_code.y) |>
+    select(glott_code.x, glott_code.y) |>
+    distinct() |>
+    mutate(type = df_name)
+
+  # Create edge list based on first three characters of style
+  edge_list_first_three <- df |>
+    mutate(first_three = substr(style, 1, 3)) |>
+    inner_join(., ., by = "first_three") |>
+    filter(glott_code.x != glott_code.y) |>
+    select(glott_code.x, glott_code.y) |>
+    distinct() |>
+    mutate(type = df_name)
+
+  # Combine both edge lists
+  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
+    distinct()
+
+  return(combined_edge_list)
+}
+
+# Apply function to all dataframes in the list and combine results
+all_edge_lists <- bind_rows(
+  lapply(names(data_list), function(df_name) {
+    create_edge_list(data_list[[df_name]], df_name)
+  })
+)
+
+```
+
+```{r}
+# Load necessary libraries
+library(dplyr)
+library(tidyr)
+
+# Subset the data
+data_list_2 <- purrr::map(data_list, ~ {
+  # Select the columns glott_code and style and convert to tibble
+selected_df <- .x %>%
+    as_tibble() %>%   # Convert to tibble
+    select(glott_code, style) %>% # Select the relevant columns
+    drop_na(style)
+
+  return(selected_df)
+})
+```
+
+```{r}
+
+# Function to create edge list for a dataframe
+create_edge_list <- function(df, df_name) {
+  # Create edge list based on first letter of style
+  df_first_letter <- df |>
+    mutate(first_letter = substr(style, 1, 1))
+  edge_list_first_letter <- df_first_letter |>
+    inner_join(df_first_letter, by = "first_letter") |>
+    filter(glott_code.x != glott_code.y) |>
+    select(glott_code.x, glott_code.y) |>
+    distinct() |>
+    mutate(type = df_name, weight = 1)
+
+  # Create edge list based on first three characters of style
+  df_first_three <- df |>
+    mutate(first_three = substr(style, 1, 3))
+  edge_list_first_three <- df_first_three |>
+    inner_join(df_first_three, by = "first_three") |>
+    filter(glott_code.x != glott_code.y) |>
+    select(glott_code.x, glott_code.y) |>
+    distinct() |>
+    mutate(type = df_name, weight = 1)
+
+  # Combine both edge lists
+  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
+    distinct()
+
+  return(combined_edge_list)
+}
+
+# Apply function to all dataframes in the list and combine results
+all_edge_lists <- bind_rows(
+  lapply(names(data_list_2), function(df_name) {
+    create_edge_list(data_list_2[[df_name]], df_name)
+  })
+)
+
+# Summarize duplicates and calculate the weight
+all_edge_lists_summarized <- all_edge_lists |>
+  group_by(glott_code.x, glott_code.y, type) |>
+  summarize(weight = sum(weight), .groups = 'drop')
+```
+
+Now, add the edge list and the node list to a tidygraph network
+
+```{r}
+library(tidygraph)
+
+g <- tbl_graph(edges = all_edge_lists_summarized, directed = FALSE)
+
+plot(g)
 ```

### 2025-03-26 analysis/paper/paper.qmd
change data cleaning code

@@
     df <- df |> 
-      dplyr::distinct(glott_code, form, .keep_all = TRUE)
+      dplyr::distinct(glott_code, style, .keep_all = TRUE)
     return(df)
@@
 #| fig-cap: "fonio word forms"
- create_leaflet_map(data_list[[1]], background = "bw")
+# create_leaflet_map(data_list$iron, background = "bw")
 ```
@@
 #| eval: false
-save_maps(data_list)
+# save_maps(data_list)
 ```

### 2025-03-27 analysis/paper/paper.qmd
add archaeological points

@@
 
+### Add further layers to the map
+Get the archaeological data layers
+```{r}
+# Create a vector of file paths
+points <- here::here("analysis", "data", "raw_data", c("gold_points.csv", 
+                                                     "fonio_points.csv",
+                                                     "millet_points.csv",
+                                                     "salt_points.csv",
+                                                     "iron_points.csv"))
+
+# read the csv files into a list of data frames
+point_list <- purrr::map(points, readr::read_csv2)
+
+# make sf geometry
+test <- point_list[[1]] |> sf::st_as_sf(coords = c("lon","lat"))
+
+# set crs
+sf::st_crs(test) <- 4326
+```
+
+function to add further layers
+```{r}
+add_sf_to_leaflet_map <- function(leaflet_map, new_data_sf, label_column) {
+
+  # Ensure the new input data is an sf object and contains the required columns
+  if (!inherits(new_data_sf, "sf")) {
+    stop("The new input data must be an sf object.")
+  }
+  if (!label_column %in% colnames(new_data_sf)) {
+    stop(paste("The column", label_column, "is not found in the new data."))
+  }
+
+  # Define a color for the new points
+  new_point_color <- "black"
+
+  # Add the new sf object as a point layer to the existing leaflet map
+  leaflet_map <- leaflet_map |>
+    leaflet::addCircleMarkers(
+      data = new_data_sf,
+      lng = ~sf::st_coordinates(geometry)[, 1],
+      lat = ~sf::st_coordinates(geometry)[, 2],
+      color = new_point_color,
+      radius = 2,
+      stroke = FALSE, fillOpacity = 1,
+      label = purrr::map(
+        glue::glue("<span style='color:{new_point_color}; font-weight: normal; font-size: 10px;'>{as.character(new_data_sf[[label_column]])}</span>"),
+        htmltools::HTML
+      ),
+      labelOptions = leaflet::labelOptions(
+        noHide = TRUE,
+        direction = 'top',
+        textOnly = TRUE
+      )
+    )
+
+  # Return the updated map object
+  return(leaflet_map)
+}
+```
+
+```{r}
+#run mapping function
+create_leaflet_map(data_list$gold, background = "bw") |>
+
+#add point layer
+add_sf_to_leaflet_map(test, label_column = "site")
+```
+
 ## Make static maps (for printing etc)
@@
 
-  # Get the background map using maptiles package
-  background_map <- maptiles::get_tiles(data_sf, provider = "Esri.WorldGrayCanvas", crop = TRUE)
-
   # Increase the boundaries of the plot
@@
   y_range <- bbox["ymax"] - bbox["ymin"]
-  bbox["xmin"] <- bbox["xmin"] - 50 * x_range
-  bbox["xmax"] <- bbox["xmax"] + 50 * x_range
-  bbox["ymin"] <- bbox["ymin"] - 50 * y_range
-  bbox["ymax"] <- bbox["ymax"] + 50 * y_range
+  bbox["xmin"] <- bbox["xmin"] - 0.5 * x_range
+  bbox["xmax"] <- bbox["xmax"] + 0.5 * x_range
+  bbox["ymin"] <- bbox["ymin"] - 0.5 * y_range
+  bbox["ymax"] <- bbox["ymax"] + 0.5 * y_range
+  attr(bbox, "class") <- "bbox"
+
+  # Get the background map using maptiles package without cropping
+  background_map <- maptiles::get_tiles(data_sf, provider = "Esri.WorldGrayCanvas", crop = FALSE)
 
@@
   tmap::tmap_mode("plot")
-  map_plot <- tmap::tm_shape(background_map) +
+  map_plot <- tmap::tm_shape(background_map, bbox = bbox) +
     tmap::tm_rgb() +
     tmap::tm_shape(data_sf) +
-    tmap::tm_text(text = form_column, col = style_column, palette = dark_palette) +
+    tmap::tm_text(text = form_column, col = style_column, 
+                  col.scale = tmap::tm_scale(values = dark_palette)) +
     tmap::tm_layout(title = plot_title, legend.show = FALSE, bbox = bbox)

### 2025-05-06 analysis/paper/loan word maps.qmd
tidying repo

--- 2025-03-27 analysis/paper/paper.qmd
@@
 ---
-title: "Make maps of loan words"
+title: "Make interactive loan word maps"
 author:
   - Henning Schreiber:
-      correspondence: "yes"
-      email: janedoe@fosg.org
-      orcid: 0000-0003-1689-0557
+      orcid: 0000-0001-9576-8036
       institute: uhh
@@
   - Nikolas Gestrich:
+      correspondence: "yes"
+      email: gestrich@uni-frankfurt.de
       institute: fi
-      orcid: 0000-0003-1689-0558
+      orcid: 0000-0001-6376-1972
 
@@
   - uhh:
-      name: Universität Hamburg
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fi: Frobenius Institute
+      name: Asien-Afrika-Institut, Universität Hamburg
+      address: Mittelweg 177, 20148 Hamburg
+  - fi: 
+      name: Frobenius Institute
+      address: Norbert-Wollheim-Platz 1, 60323 Frankfurt am Main, Gemany
 title-block-published: "Last updated"  
@@
 Modify the list of .csv files to suit your interests.
-If you want to upload new files, place these in the directory /analysis/data/raw_data.Make sure any new data has a glottolog language code in a varaible named "glott_code"
+If you want to upload new files, place these in the directory /analysis/data/raw_data.Make sure any new data has a glottolog language code in a variable named "glott_code"
 
@@
 # make sf geometry
-test <- point_list[[1]] |> sf::st_as_sf(coords = c("lon","lat"))
+points_sf <- point_list[[1]] |> sf::st_as_sf(coords = c("lon","lat"))
 
 # set crs
-sf::st_crs(test) <- 4326
+sf::st_crs(points_sf) <- 4326
 ```
@@
 #add point layer
-add_sf_to_leaflet_map(test, label_column = "site")
-```
-
-## Make static maps (for printing etc)
-
-Function to make static plots
-
-```{r}
-create_static_map <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {
-
-  # Ensure the input data is an sf object and contains the required columns
-  if (!inherits(data_sf, "sf")) {
-    stop("The input data must be an sf object.")
-  }
-  if (!style_column %in% colnames(data_sf)) {
-    stop(paste("The column", style_column, "is not found in the data."))
-  }
-  if (!"glott_code" %in% colnames(data_sf)) {
-    stop("The column 'glott_code' is not found in the data.")
-  }
-  if (!"LanName" %in% colnames(data_sf)) {
-    stop("The column 'LanName' is not found in the data.")
-  }
-
-  # Filter out rows with NA values in the "style" column
-  data_sf <- data_sf |> dplyr::filter(!is.na(!!rlang::sym(style_column)))
-
-  # Define the color palette for the "style" column, ensuring darker colors
-  num_categories <- length(unique(data_sf[[style_column]]))
-  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)
-
-  # Create the base map with background
-  base_map <- ggplot2::ggplot() +
-    ggspatial::annotation_map_tile(tiles = "http://server.arcgisonline.com/ArcGIS/rest/services/Canvas/World_Light_Gray_Base/MapServer/tile/${z}/${y}/${x}") +
-    ggplot2::geom_sf(data = data_sf, ggplot2::aes(color = !!rlang::sym(style_column)), show.legend = FALSE) +
-    ggplot2::scale_color_manual(values = dark_palette) +
-    ggplot2::theme_minimal() +
-    ggplot2::theme(legend.position = "bottom") +
-    ggplot2::labs(color = style_column)
-
-  # Add the plot title if provided
-  if (!is.null(plot_title)) {
-    base_map <- base_map + ggplot2::ggtitle(plot_title)
-  }
-
-  # Crop the extent of the map to the data layers' extent
-  base_map <- base_map + ggplot2::coord_sf(xlim = sf::st_bbox(data_sf)[c("xmin", "xmax")],
-                                           ylim = sf::st_bbox(data_sf)[c("ymin", "ymax")])
-
-  # Plot the data as strings contained in the form_column variable
-  map_plot <- base_map +
-    ggplot2::geom_text(data = data_sf, ggplot2::aes(x = sf::st_coordinates(data_sf)[,1],
-                                                    y = sf::st_coordinates(data_sf)[,2],
-                                                    label = !!rlang::sym(form_column),
-                                                    color = !!rlang::sym(style_column)),
-                       position = ggplot2::position_jitter(width = 0.1, height = 0.1), size = 3)
-
-  # Return the ggplot object
-  return(map_plot)
-}
-```
-
-Make static plots
-
-```{r}
-create_static_map(data_list[[2]], plot_title = "fonio")
-```
-
-```{r}
-create_static_map_tmap <- function(data_sf, form_column = "form", style_column = "style", plot_title = NULL) {
-
-  # Ensure the input data is an sf object and contains the required columns
-  if (!inherits(data_sf, "sf")) {
-    stop("The input data must be an sf object.")
-  }
-  if (!style_column %in% colnames(data_sf)) {
-    stop(paste("The column", style_column, "is not found in the data."))
-  }
-  if (!"glott_code" %in% colnames(data_sf)) {
-    stop("The column 'glott_code' is not found in the data.")
-  }
-  if (!"LanName" %in% colnames(data_sf)) {
-    stop("The column 'LanName' is not found in the data.")
-  }
-
-  # Filter out rows with NA values in the "style" column
-  data_sf <- data_sf |> dplyr::filter(!is.na(!!rlang::sym(style_column)))
-
-  # Define the color palette for the "style" column, ensuring darker colors
-  num_categories <- length(unique(data_sf[[style_column]]))
-  dark_palette <- scales::brewer_pal(palette = "Dark2")(num_categories)
-
-  # Increase the boundaries of the plot
-  bbox <- sf::st_bbox(data_sf)
-  x_range <- bbox["xmax"] - bbox["xmin"]
-  y_range <- bbox["ymax"] - bbox["ymin"]
-  bbox["xmin"] <- bbox["xmin"] - 0.5 * x_range
-  bbox["xmax"] <- bbox["xmax"] + 0.5 * x_range
-  bbox["ymin"] <- bbox["ymin"] - 0.5 * y_range
-  bbox["ymax"] <- bbox["ymax"] + 0.5 * y_range
-  attr(bbox, "class") <- "bbox"
-
-  # Get the background map using maptiles package without cropping
-  background_map <- maptiles::get_tiles(data_sf, provider = "Esri.WorldGrayCanvas", crop = FALSE)
-
-  # Create the tmap object
-  tmap::tmap_mode("plot")
-  map_plot <- tmap::tm_shape(background_map, bbox = bbox) +
-    tmap::tm_rgb() +
-    tmap::tm_shape(data_sf) +
-    tmap::tm_text(text = form_column, col = style_column, 
-                  col.scale = tmap::tm_scale(values = dark_palette)) +
-    tmap::tm_layout(title = plot_title, legend.show = FALSE, bbox = bbox)
-
-  # Return the tmap object
-  return(map_plot)
-}
-```
-
-```{r}
-create_static_map_tmap(data_list[[1]])
-```
-
-## Make a network
-
-Our network will be made using languages as nodes, which come with locations. We will then use the data in the list to create edges between the nodes.
-
-```{r}
-# Load necessary libraries
-library(dplyr)
-library(tidyr)
-
-# Function to create edge list for a dataframe
-create_edge_list <- function(df, df_name) {
-  # Clean and transform the data
-  df <- df |> 
-    select(glott_code, style) |> 
-    as_tibble()
-
-  # Create edge list based on first letter of style
-  edge_list_first_letter <- df |>
-    mutate(first_letter = substr(style, 1, 1)) |>
-    inner_join(., ., by = "first_letter") |>
-    filter(glott_code.x != glott_code.y) |>
-    select(glott_code.x, glott_code.y) |>
-    distinct() |>
-    mutate(type = df_name)
-
-  # Create edge list based on first three characters of style
-  edge_list_first_three <- df |>
-    mutate(first_three = substr(style, 1, 3)) |>
-    inner_join(., ., by = "first_three") |>
-    filter(glott_code.x != glott_code.y) |>
-    select(glott_code.x, glott_code.y) |>
-    distinct() |>
-    mutate(type = df_name)
-
-  # Combine both edge lists
-  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
-    distinct()
-
-  return(combined_edge_list)
-}
-
-# Apply function to all dataframes in the list and combine results
-all_edge_lists <- bind_rows(
-  lapply(names(data_list), function(df_name) {
-    create_edge_list(data_list[[df_name]], df_name)
-  })
-)
-
-```
-
-```{r}
-# Load necessary libraries
-library(dplyr)
-library(tidyr)
-
-# Subset the data
-data_list_2 <- purrr::map(data_list, ~ {
-  # Select the columns glott_code and style and convert to tibble
-selected_df <- .x %>%
-    as_tibble() %>%   # Convert to tibble
-    select(glott_code, style) %>% # Select the relevant columns
-    drop_na(style)
-
-  return(selected_df)
-})
-```
-
-```{r}
-
-# Function to create edge list for a dataframe
-create_edge_list <- function(df, df_name) {
-  # Create edge list based on first letter of style
-  df_first_letter <- df |>
-    mutate(first_letter = substr(style, 1, 1))
-  edge_list_first_letter <- df_first_letter |>
-    inner_join(df_first_letter, by = "first_letter") |>
-    filter(glott_code.x != glott_code.y) |>
-    select(glott_code.x, glott_code.y) |>
-    distinct() |>
-    mutate(type = df_name, weight = 1)
-
-  # Create edge list based on first three characters of style
-  df_first_three <- df |>
-    mutate(first_three = substr(style, 1, 3))
-  edge_list_first_three <- df_first_three |>
-    inner_join(df_first_three, by = "first_three") |>
-    filter(glott_code.x != glott_code.y) |>
-    select(glott_code.x, glott_code.y) |>
-    distinct() |>
-    mutate(type = df_name, weight = 1)
-
-  # Combine both edge lists
-  combined_edge_list <- bind_rows(edge_list_first_letter, edge_list_first_three) |>
-    distinct()
-
-  return(combined_edge_list)
-}
-
-# Apply function to all dataframes in the list and combine results
-all_edge_lists <- bind_rows(
-  lapply(names(data_list_2), function(df_name) {
-    create_edge_list(data_list_2[[df_name]], df_name)
-  })
-)
-
-# Summarize duplicates and calculate the weight
-all_edge_lists_summarized <- all_edge_lists |>
-  group_by(glott_code.x, glott_code.y, type) |>
-  summarize(weight = sum(weight), .groups = 'drop')
-```
-
-Now, add the edge list and the node list to a tidygraph network
-
-```{r}
-library(tidygraph)
-
-g <- tbl_graph(edges = all_edge_lists_summarized, directed = FALSE)
-
-plot(g)
+add_sf_to_leaflet_map(points_sf, label_column = "site")
 ```

### 2025-05-06 analysis/paper/loan word maps.qmd
tidy repo

@@
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  West Africa; trade; historical linguistics
 highlights: |
-  These are the highlights. 
+  Maps of loanwords to illustrate historical trade pattern in West Africa
 prefer-html: true
@@
 
+## Retrieve and clean data
 In order to have location data, we read in the African languages, their glottoglog code, and their location from the glottolog database (@hammarstromGlottolog2024).
@@
 #| fig-cap: "fonio word forms"
-# create_leaflet_map(data_list$iron, background = "bw")
+# create_leaflet_map(data_list$iron, background = "bw") #uncomment to run
 ```
@@
 #| eval: false
-# save_maps(data_list)
+# save_maps(data_list)  #uncomment to run
 ```
@@
 #run mapping function
-create_leaflet_map(data_list$gold, background = "bw") |>
+#create_leaflet_map(data_list$gold, background = "bw") |> #uncomment to run
 
 #add point layer
-add_sf_to_leaflet_map(points_sf, label_column = "site")
+#add_sf_to_leaflet_map(points_sf, label_column = "site") #uncomment to run
 ```

### 2025-05-14 analysis/paper/loan word maps.qmd
network analysis

@@
 
+## Networks
+Next, we will create networks from this data, in order to be able to analyse and visualise connections.The dataset of languages is the nodelist, the edges are made on the basis of shared forms.
+First, the edge list is made.
+Fir this, we need more detailed classifications than for the maps above, so we need to read the data in again without simplifying the classifications.
+
+```{r prepare data for network}
+# read classified loan word data
+
+# read the csv files into a list of data frames
+net_data_list <- purrr::map(objs, readr::read_csv)
+
+# Assign names
+names(net_data_list) <- c("gold", "fonio", "millet", "salt", "iron")
+
+# Function to prepare each dataset for making it into an edge list
+prepare_data <- function(data) {
+  data |> 
+    # Step 1: Split the 'style' column into hierarchical levels
+    dplyr::mutate(
+      cl1 = stringr::str_extract(style, "^[A-Z]"),                    # First level
+      cl2 = stringr::str_extract(style, "^[A-Z]\\.\\d+"),             # Second level
+      cl3 = stringr::str_extract(style, "^[A-Z]\\.\\d+\\.\\w+")       # Third level
+    ) |> 
+    # Step 2: Replace "X" with NA in cl1, cl2, and cl3
+    dplyr::mutate(
+      cl1 = dplyr::if_else(cl1 == "X", NA_character_, cl1),
+      cl2 = dplyr::if_else(cl2 == "X", NA_character_, cl2),
+      cl3 = dplyr::if_else(cl3 == "X", NA_character_, cl3)
+    )
+}
+
+# Apply the function to each dataset in the list
+net_data_list <- lapply(net_data_list, prepare_data) 
+```
+
+```{r edge list function}
+# Function to generate edge list
+generate_edgelist <- function(data) {
+  # Step 1: Create pairwise comparisons of glott_code values
+  edgelist <- expand.grid(glott_code_1 = data$glott_code, 
+                          glott_code_2 = data$glott_code,
+                          stringsAsFactors = FALSE) |>
+    # Remove self-loops
+    dplyr::filter(glott_code_1 != glott_code_2) |>
+    # Add corresponding cl1, cl2, cl3 for both nodes
+    dplyr::left_join(data, by = c("glott_code_1" = "glott_code")) |>
+    dplyr::rename(cl1_1 = cl1, cl2_1 = cl2, cl3_1 = cl3) |>
+    dplyr::left_join(data, by = c("glott_code_2" = "glott_code")) |>
+    dplyr::rename(cl1_2 = cl1, cl2_2 = cl2, cl3_2 = cl3) |>
+    # Calculate weights based on equivalence
+    dplyr::mutate(
+      weight = dplyr::case_when(
+        cl1_1 == cl1_2 & cl2_1 == cl2_2 & cl3_1 == cl3_2 ~ 3,  # Match in cl1, cl2, and cl3
+        cl1_1 == cl1_2 & cl2_1 == cl2_2 ~ 2,                  # Match in cl1 and cl2
+        cl1_1 == cl1_2 ~ 1,                                   # Match in cl1 only
+        TRUE ~ 0                                              # No match
+      )
+    ) |>
+    # Filter out pairs with weight = 0 (no equivalence)
+    dplyr::filter(weight > 0) |>
+    dplyr::select(glott_code_1, glott_code_2, weight)
+
+  # Step 2: Reduce the edgelist to retain only one edge per pair with the highest weight
+  reduced_edgelist <- edgelist |>
+    # Ensure undirected edges by sorting the glott_code pair
+    dplyr::mutate(
+      node_1 = pmin(glott_code_1, glott_code_2),
+      node_2 = pmax(glott_code_1, glott_code_2)
+    ) |>
+    # Group by the unique undirected pair
+    dplyr::group_by(node_1, node_2) |>
+    # Retain only the edge with the maximum weight
+    dplyr::summarise(weight = max(weight), .groups = "drop") |>
+    # Rename columns back to glott_code_1 and glott_code_2 for clarity
+    dplyr::rename(glott_code_1 = node_1, glott_code_2 = node_2)
+
+  return(reduced_edgelist)
+}
+```
+
+```{r create edge lists}
+edgelists <- lapply(net_data_list, generate_edgelist)
+```
+
+now, we read this into tidygraph network objects. First we take the language list and convert it into a usable edge list, and then make both into a network.
+```{r}
+# Convert langloc into a tibble with latitude and longitude fields
+node_list <- langloc |>
+  dplyr::mutate(
+    latitude = sf::st_coordinates(geometry)[, 2],
+    longitude = sf::st_coordinates(geometry)[, 1]
+  ) |>
+  tibble::as_tibble() |> 
+  dplyr::select(glott_code, LanName, latitude, longitude) |>
+
+#add some missing values
+ tibble::add_row(glott_code = "dogo1299", 
+                             LanName = "Dogon", 
+                             latitude = 14.43, 
+                             longitude = -3.49 )
+
+# Create a list of tidygraph objects using the filtered node list
+graphs <- lapply(edgelists, function(edges) {
+  # Filter the node list to include only nodes present in the edge list
+  filtered_nodes <- node_list |>
+    dplyr::filter(glott_code %in% c(edges$glott_code_1, edges$glott_code_2))
+
+  # Create the tidygraph object
+  tidygraph::tbl_graph(
+    nodes = filtered_nodes,
+    edges = edges,
+    directed = FALSE
+  )
+})
+
+```
+
+```{r graph plotting functions}
+# Function to plot a tidygraph object with geographic layout
+plot_geo_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight", label = "LanName", title = "title") {
+  # Ensure that the graph has the required attributes
+  if (!latitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Latitude column", latitude, "not found in node attributes."))
+  }
+  if (!longitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Longitude column", longitude, "not found in node attributes."))
+  }
+  if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
+    stop(paste("Weight column", weight, "not found in edge attributes."))
+  }
+  if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Label column", label, "not found in node attributes."))
+  }
+
+  # Plot the tidygraph object
+  ggraph::ggraph(graph, layout = "manual", x = !!rlang::sym(longitude), y = !!rlang::sym(latitude)) +
+    ggraph::geom_edge_link(
+      ggplot2::aes(width = !!rlang::sym(weight), alpha = !!rlang::sym(weight)), 
+      color = "gray"
+    ) +
+    ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
+    ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
+    ggraph::geom_node_point(size = 5, color = "blue") +
+    ggraph::geom_node_text(
+      ggplot2::aes(label = !!rlang::sym(label)), 
+      repel = TRUE, 
+      size = 3, 
+      color = "black"
+    ) +
+    ggplot2::labs(title = title) +
+    ggraph::theme_graph() +
+    ggplot2::theme(
+      legend.position = "none"
+    )
+}
+
+# Function to plot a tidygraph object with force-directed layout
+plot_fr_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight", label = "LanName", title = "title") {
+  # Ensure that the graph has the required attributes
+  if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
+    stop(paste("Weight column", weight, "not found in edge attributes."))
+  }
+  if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Label column", label, "not found in node attributes."))
+  }
+
+  # Plot the tidygraph object
+  ggraph::ggraph(graph, layout = "fr") +
+    ggraph::geom_edge_link(
+      ggplot2::aes(width = !!rlang::sym(weight), alpha = !!rlang::sym(weight)), 
+      color = "gray"
+    ) +
+    ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
+    ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
+    ggraph::geom_node_point(size = 5, color = "blue") +
+    ggraph::geom_node_text(
+      ggplot2::aes(label = !!rlang::sym(label)), 
+      repel = TRUE, 
+      size = 3, 
+      color = "black"
+    ) +
+    ggplot2::labs(title = title) +
+    ggraph::theme_graph() +
+    ggplot2::theme(
+      legend.position = "none"
+    )
+}
+
+# Function to plot a tidygraph object with geographic layout and louvain communities
+plot_comm_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight", label = "LanName", title = "title", subtitle = NULL, community = "louvain") {
+  # Ensure that the graph has the required attributes
+  if (!latitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Latitude column", latitude, "not found in node attributes."))
+  }
+  if (!longitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Longitude column", longitude, "not found in node attributes."))
+  }
+  if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
+    stop(paste("Weight column", weight, "not found in edge attributes."))
+  }
+  if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Label column", label, "not found in node attributes."))
+  }
+   if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+    stop(paste("Label column", community, "not found in node attributes."))
+  }
+
+  # Plot the tidygraph object
+  ggraph::ggraph(graph, layout = "manual", x = !!rlang::sym(longitude), y = !!rlang::sym(latitude)) +
+    ggraph::geom_edge_link(
+      ggplot2::aes(width = !!rlang::sym(weight), alpha = !!rlang::sym(weight)), 
+      color = "gray"
+    ) +
+    ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
+    ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
+    ggraph::geom_node_point(aes(colour = !!rlang::sym(community)),size = 5) +
+    ggraph::geom_node_text(
+      ggplot2::aes(label = !!rlang::sym(label)), 
+      repel = TRUE, 
+      size = 3, 
+      color = "black"
+    ) +
+    ggplot2::labs(title = title, subtitle = subtitle) +
+    ggraph::theme_graph() +
+    ggplot2::theme(
+      legend.position = "none"
+    )
+}
+
+```
+
+```{r}
+plot_geo_graph(graphs$gold, title = "gold")
+```
+
+Improve the gold graph:
+remove Wolof. It has the same form as the eastern Group, but its ouro form is probably portuguese influenced, whereas the lower Niger is not.
+Retain only stronger ties.
+
+```{r}
+library(tidygraph)
+graphs$gold |> 
+  mutate(louvain = group_louvain(weights = weight)) |> 
+  mutate(louvain = as.character(louvain)) |> 
+  activate(edges) |> 
+  filter(weight > 1) |> 
+  activate(nodes) |>
+  filter(!node_is_isolated()) |>
+
+  plot_comm_graph(community = "louvain", title = "gold", subtitle = "Louvain network communities")
+
+```
+
+```{r}
+graphs$salt |> 
+  mutate(louvain = group_louvain(weights = weight)) |> 
+  mutate(louvain = as.character(louvain)) |> 
+  activate(edges) |> 
+  filter(weight > 1) |> 
+  activate(nodes) |>
+  filter(!node_is_isolated()) |>
+
+  plot_comm_graph(community = "louvain", title = "salt", subtitle = "Louvain network communities")
+```
+
+Create a graph each with betweenness centrality
+```{r betweenness graphs}
+library(tidygraph)
+library(ggraph)
+library(dplyr)
+
+# Iterate over each "type" in the edge list and create a graph
+edge_list_summary |>
+  dplyr::group_split(type) |>
+  purrr::walk(\(edges_for_type) {
+    # Get the type name
+    type_name <- unique(edges_for_type$type)
+
+    # Create a filtered tidygraph object for this type
+    graph_for_type <- tidygraph::tbl_graph(
+      nodes = node_list,
+      edges = edges_for_type,
+      directed = FALSE
+    ) |>
+      tidygraph::filter(!tidygraph::node_is_isolated()) # Remove isolated nodes
+
+    # Compute betweenness centrality for node sizes
+    graph_for_type <- graph_for_type |>
+      tidygraph::mutate(betweenness = tidygraph::centrality_betweenness())
+
+    # Plot the graph
+    plot <- ggraph::ggraph(graph_for_type, layout = "manual", x = longitude, y = latitude) +
+      ggraph::geom_edge_link(aes(width = weight), edge_colour = "grey") +
+      ggraph::geom_node_point(aes(size = betweenness), colour = "blue") +
+      ggraph::geom_node_text(aes(label = LanName), repel = TRUE, size = 3) +
+      ggplot2::theme_void() +
+      ggplot2::labs(
+        title = paste("Graph for Type:", type_name),
+        size = "Betweenness",
+        width = "Edge Weight"
+      ) +
+      ggplot2::scale_size_continuous(range = c(2, 10)) +
+      ggraph::scale_edge_width_continuous(range = c(0.5, 3)) +
+      ggraph::theme_graph(base_size = 12)
+
+    print(plot)
+  })
+```
+Create a graph each with eigenvector centrality
+```{r betweenness graphs}
+library(tidygraph)
+library(ggraph)
+library(dplyr)
+
+# Iterate over each "type" in the edge list and create a graph
+edge_list_summary |>
+  dplyr::group_split(type) |>
+  purrr::walk(\(edges_for_type) {
+    # Get the type name
+    type_name <- unique(edges_for_type$type)
+
+    # Create a filtered tidygraph object for this type
+    graph_for_type <- tidygraph::tbl_graph(
+      nodes = node_list,
+      edges = edges_for_type,
+      directed = FALSE
+    ) |>
+      tidygraph::filter(!tidygraph::node_is_isolated()) # Remove isolated nodes
+
+    # Compute betweenness centrality for node sizes
+    graph_for_type <- graph_for_type |>
+      tidygraph::mutate(eigenvector = tidygraph::centrality_eigen())
+
+    # Plot the graph
+    plot <- ggraph::ggraph(graph_for_type, layout = "manual", x = longitude, y = latitude) +
+      ggraph::geom_edge_link(aes(width = weight), edge_colour = "grey") +
+      ggraph::geom_node_point(aes(size = eigenvector), colour = "purple") +
+      ggraph::geom_node_text(aes(label = LanName), repel = TRUE, size = 3) +
+      ggplot2::theme_void() +
+      ggplot2::labs(
+        title = paste("Graph for Type:", type_name),
+        size = "Eigenvector",
+        width = "Edge Weight"
+      ) +
+      ggplot2::scale_size_continuous(range = c(2, 10)) +
+      ggraph::scale_edge_width_continuous(range = c(0.5, 3)) +
+      ggraph::theme_graph(base_size = 12)
+
+    print(plot)
+  })
+```
+
+Save a graph for each of the trade items
+```{r}
+# Create a list of tidygraph objects, one for each "type"
+graph_list <- edge_list_summary |>
+  dplyr::group_split(type) |>
+  purrr::map(\(edges_for_type) {
+    # Get the type name
+    type_name <- unique(edges_for_type$type)
+
+    # Create a filtered tidygraph object for this type
+    graph_for_type <- tidygraph::tbl_graph(
+      nodes = node_list,
+      edges = edges_for_type,
+      directed = FALSE
+    ) |>
+      tidygraph::filter(!tidygraph::node_is_isolated()) |> # Remove isolated nodes
+      tidygraph::mutate(betweenness = tidygraph::centrality_betweenness()) # Add betweenness centrality
+
+    # Return the graph, named with the type
+    graph_for_type
+  }) |>
+  purrr::set_names(unique(edge_list_summary$type))
+
+# Save the list of graphs for further processing
+save(graph_list, file = "graph_list.RData")
+
+```
+
+```{r}
+
+```
+
+```{r}
+#knock out WOlof from the gold network
+graph |> 
+  activate(edges) |> 
+  filter(type == "gold") |> 
+  activate(nodes) |> 
+  filter(LanName != "Wolof") |>
+  filter(!node_is_isolated()) |> 
+  plot_graph()
+```
+
+The graphs need to be sparsified now, as they are relatively useless as they are.
+```{r}
+library(backbone)
+
+disp <- disparity(graph_list$gold, alpha = 0.5)
+plot_graph(disp)
+
+```
+
+```{r}
+# Find glott_code values in edge lists that are not in the node_list
+find_missing_glott_codes <- function(edgelists, node_list) {
+  # Combine all unique glott_code values from the edge lists
+  all_glott_codes <- lapply(edgelists, function(edges) {
+    unique(c(edges$glott_code_1, edges$glott_code_2))
+  }) |>
+    unlist() |>  # Flatten into a single vector
+    unique()     # Remove duplicates
+
+  # Find glott_codes not present in the node_list
+  missing_glott_codes <- setdiff(all_glott_codes, node_list$glott_code)
+
+  return(missing_glott_codes)
+}
+
+# Apply the function
+missing_glott_codes <- find_missing_glott_codes(edgelists, node_list)
+
+# Display the missing glott_code values
+print(missing_glott_codes)
+
+```
+
 <!-- The following line inserts a page break  -->

### 2025-06-02 analysis/paper/loan word maps.qmd
add network analysis

@@
 
+```
+
+### gold
+
+```{r}
+library(tidygraph) # read in tidygraph library for less typing during exploration
+```
+
+```{r}
+plot_geo_graph(graphs$gold, title = "gold")
+```
+
+Improve the gold graph:
+remove Wolof. It has the same form as the eastern Group, but its ouro form is probably portuguese influenced, whereas the lower Niger is not.
+Retain only stronger ties.
+
+```{r louvain community plot function}
+plot_louvain_graph <- function(item) {
+  graph <- graphs[[item]] |>
+    tidygraph::mutate(louvain = tidygraph::group_louvain(weights = weight)) |>
+    tidygraph::mutate(louvain = as.character(louvain)) |>
+    tidygraph::activate("edges") |>
+    tidygraph::filter(weight > 1) |>
+    tidygraph::activate("nodes") |>
+    tidygraph::filter(!tidygraph::node_is_isolated())
+
       # Function to plot a tidygraph object with geographic layout and louvain communities
@@
         ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
-    ggraph::geom_node_point(aes(colour = !!rlang::sym(community)),size = 5) +
+        ggraph::geom_node_point(ggplot2::aes(colour = !!rlang::sym(community)),size = 5) +
         ggraph::geom_node_text(
@@
 
+  plot_comm_graph(
+    graph,
+    community = "louvain",
+    title = item,
+    subtitle = "Louvain network communities"
+  )
+}
 ```
@@
 ```{r}
-plot_geo_graph(graphs$gold, title = "gold")
+items <- names(net_data_list)
+
+lapply(items, plot_louvain_graph)
 ```
 
-Improve the gold graph:
-remove Wolof. It has the same form as the eastern Group, but its ouro form is probably portuguese influenced, whereas the lower Niger is not.
-Retain only stronger ties.
+Can we sparsify using random walks or similar traversing techniques? Trade networks in economics.
+
+### Fonio
+```{r }
+plot_fr_graph(graphs$fonio, title = "fonio")
+```
+
+This graph is somewhat perplexing. What does the connection between Bangime (wīī-bɛ̀) and Igo (ā-wúbī,volta-Congo langue in southern Togo) mean? This is a question that needs further answering. 
+For now, we can remove this pair, and concentrate on the rest.
 
 ```{r}
-library(tidygraph)
-graphs$gold |> 
-  mutate(louvain = group_louvain(weights = weight)) |> 
-  mutate(louvain = as.character(louvain)) |> 
-  activate(edges) |> 
-  filter(weight > 1) |> 
+graphs$fonio |> 
   activate(nodes) |> 
-  filter(!node_is_isolated()) |>
+  filter(!LanName %in% c("Bangime", "Igo")) |> 
+  plot_fr_graph(title = "fonio")
+```
 
-  plot_comm_graph(community = "louvain", title = "gold", subtitle = "Louvain network communities")
+We can now see that there are outliers at either end.
+One is composed of a large number of Songhai variants, including Eastern Songhai Koyraboro, Humburi, and Zarma, Northern Songhai Tadaksahak, and the pastoralist languages Tuareg and Maasina Fulfulde.
+It is very curious that Zarma and Dendi, which are part of the same subdivision of Eastern Songhai, fall on opposite ends of this graph.
+We will remove both ends now to better see the centre, which appears to have three groups.
+
+```{r}
+graphs$fonio |> 
+  activate(nodes) |> 
+  filter(!LanName %in% c("Bangime", 
+                         "Igo",
+                         "Tuareg",
+                         "Zarma",
+                         "Tadaksahak",
+                         "Koyraboro Senni Songhai",
+                         "Humburi Senni Songhay",
+                         "Maasina Fulfulde",
+                         "Dendi (Benin)")) |> 
+  plot_fr_graph(title = "fonio")
+```
+
+The resulting two groups, with a group of connecting languages between, are a geographic mess.
+Often, language groups fall on either end.
+This makes me think that more detailed analysis is necessary here.
+Are we picking up generics and specifics and lumping them together?
+There is a lot more work necessary on the vocabulary in order to make this happen.
+The problem is that this is not such a well-defined item as, for instance, gold is.
+
+### Millet
+
+```{r}
+plot_fr_graph(graphs$millet, title = "millet")
+```
+
+Despite the fact that we did not note millet as a trade item, this is a super interesting graph. Let's begin with the outliers: Wan (SE Mandé, CdI) has an unconnected form (fen), and Dagaari Dioula
+
+```{r}
+graphs$millet |> 
+  activate(nodes) |> 
+  filter(!LanName %in% c("Mwan", "Toura (Côte d'Ivoire)", "Baoulé", "Alladian", "Dagaari Dioula", "Wan")) |> 
+  plot_fr_graph(title = "millet")
+```
+
+## Finding trade routes in the networks
+
+### Create a cost variable
+This combines weight and geographic distance.
+```{r}
+graph <- graphs$gold
+
+library(geosphere)
+
+# Extract coordinates
+coords <- as.data.frame(tidygraph::as_tibble(graph, active = "nodes"))
+
+# Compute geographic distance for each edge
+edge_info <- tidygraph::as_tibble(graph, active = "edges") |>
+  dplyr::mutate(
+    lat_from = coords$latitude[from],
+    lon_from = coords$longitude[from],
+    lat_to = coords$latitude[to],
+    lon_to = coords$longitude[to],
+    geo_dist = geosphere::distHaversine(cbind(lon_from, lat_from), cbind(lon_to, lat_to)),
+    cost = geo_dist / weight
+  )
+
+# Add back to graph
+graph <- tidygraph::activate(graph, "edges") |>
+  dplyr::mutate(
+    geo_dist = edge_info$geo_dist,
+    cost = edge_info$cost
+  )
+```
+
+### Via edge betweenness
+Goal: Find edges that are most frequently used in shortest paths — "key trade routes"
+```{r}
+graph <- graph |> 
+  tidygraph::activate(edges) |> 
+  tidygraph::mutate(inv_cost = ifelse(cost == 0, Inf, 1 / cost)) |> # cost variable needs to be inverted
+  tidygraph::mutate(edge_betweenness = centrality_edge_betweenness(weights = inv_cost))
+
+# Activate edges
+edge_data <- tidygraph::activate(graph, "edges") |> 
+  tidygraph::as_tibble()
+
+# Summary statistics
+summary(edge_data$edge_betweenness)
+
+# Histogram
+hist(edge_data$edge_betweenness, 
+     breaks = 30, 
+     main = "Distribution of Edge Betweenness", 
+     xlab = "Edge Betweenness", 
+     col = "steelblue", 
+     border = "white")
 
@@
 ```{r}
-graphs$salt |> 
-  mutate(louvain = group_louvain(weights = weight)) |> 
-  mutate(louvain = as.character(louvain)) |> 
-  activate(edges) |> 
-  filter(weight > 1) |> 
-  activate(nodes) |>
-  filter(!node_is_isolated()) |>
+plot_edgecost_graph <- function(input_graph, cost, cutoff, item) {
+  graph <- input_graph |>
+    tidygraph::activate("nodes") |> 
+    tidygraph::mutate(louvain = tidygraph::group_louvain(weights = weight)) |>
+    tidygraph::mutate(louvain = as.character(louvain)) |>
+    tidygraph::activate("edges") |> 
+    tidygraph::filter(edge_betweenness>cutoff) |> 
+    tidygraph::activate("nodes") |> 
+    tidygraph::filter(!tidygraph::node_is_isolated())
 
-  plot_comm_graph(community = "louvain", title = "salt", subtitle = "Louvain network communities")
-```
+      # Function to plot a tidygraph object with geographic layout and louvain communities
+    plot_comm_graph <- function(graph, latitude = "latitude", longitude = "longitude", weight = "weight",     label = "LanName", title = "title", subtitle = NULL, community = "louvain") {
+      # Ensure that the graph has the required attributes
+      if (!latitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+        stop(paste("Latitude column", latitude, "not found in node attributes."))
+      }
+      if (!longitude %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+        stop(paste("Longitude column", longitude, "not found in node attributes."))
+      }
+      if (!weight %in% colnames(tidygraph::as_tibble(graph, active = "edges"))) {
+        stop(paste("Weight column", weight, "not found in edge attributes."))
+      }
+      if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+        stop(paste("Label column", label, "not found in node attributes."))
+      }
+       if (!label %in% colnames(tidygraph::as_tibble(graph, active = "nodes"))) {
+        stop(paste("Label column", community, "not found in node attributes."))
+      }
 
-Create a graph each with betweenness centrality
-```{r betweenness graphs}
-library(tidygraph)
-library(ggraph)
-library(dplyr)
+      # Plot the tidygraph object
+      ggraph::ggraph(graph, layout = "manual", x = !!rlang::sym(longitude), y = !!rlang::sym(latitude)) +
+        ggraph::geom_edge_link(
+          ggplot2::aes(width = !!rlang::sym(cost), alpha = !!rlang::sym(cost)), 
+          color = "gray"
+        ) +
+        ggraph::scale_edge_alpha_continuous(range = c(0.6, 1)) + # Map weight to opacity
+        ggraph::scale_edge_width_continuous(range = c(0.6, 1)) + # Map weight to width
+        ggraph::geom_node_point(ggplot2::aes(colour = !!rlang::sym(community)),size = 5) +
+        ggraph::geom_node_text(
+          ggplot2::aes(label = !!rlang::sym(label)), 
+          repel = TRUE, 
+          size = 3, 
+          color = "black"
+        ) +
+        ggplot2::labs(title = title, subtitle = subtitle) +
+        ggraph::theme_graph() +
+        ggplot2::theme(
+          legend.position = "none"
+        )
+    }
 
-# Iterate over each "type" in the edge list and create a graph
-edge_list_summary |>
-  dplyr::group_split(type) |>
-  purrr::walk(\(edges_for_type) {
-    # Get the type name
-    type_name <- unique(edges_for_type$type)
-
-    # Create a filtered tidygraph object for this type
-    graph_for_type <- tidygraph::tbl_graph(
-      nodes = node_list,
-      edges = edges_for_type,
-      directed = FALSE
-    ) |>
-      tidygraph::filter(!tidygraph::node_is_isolated()) # Remove isolated nodes
-
-    # Compute betweenness centrality for node sizes
-    graph_for_type <- graph_for_type |>
-      tidygraph::mutate(betweenness = tidygraph::centrality_betweenness())
-
-    # Plot the graph
-    plot <- ggraph::ggraph(graph_for_type, layout = "manual", x = longitude, y = latitude) +
-      ggraph::geom_edge_link(aes(width = weight), edge_colour = "grey") +
-      ggraph::geom_node_point(aes(size = betweenness), colour = "blue") +
-      ggraph::geom_node_text(aes(label = LanName), repel = TRUE, size = 3) +
-      ggplot2::theme_void() +
-      ggplot2::labs(
-        title = paste("Graph for Type:", type_name),
-        size = "Betweenness",
-        width = "Edge Weight"
-      ) +
-      ggplot2::scale_size_continuous(range = c(2, 10)) +
-      ggraph::scale_edge_width_continuous(range = c(0.5, 3)) +
-      ggraph::theme_graph(base_size = 12)
-
-    print(plot)
-  })
-```
-Create a graph each with eigenvector centrality
-```{r betweenness graphs}
-library(tidygraph)
-library(ggraph)
-library(dplyr)
-
-# Iterate over each "type" in the edge list and create a graph
-edge_list_summary |>
-  dplyr::group_split(type) |>
-  purrr::walk(\(edges_for_type) {
-    # Get the type name
-    type_name <- unique(edges_for_type$type)
-
-    # Create a filtered tidygraph object for this type
-    graph_for_type <- tidygraph::tbl_graph(
-      nodes = node_list,
-      edges = edges_for_type,
-      directed = FALSE
-    ) |>
-      tidygraph::filter(!tidygraph::node_is_isolated()) # Remove isolated nodes
-
-    # Compute betweenness centrality for node sizes
-    graph_for_type <- graph_for_type |>
-      tidygraph::mutate(eigenvector = tidygraph::centrality_eigen())
-
-    # Plot the graph
-    plot <- ggraph::ggraph(graph_for_type, layout = "manual", x = longitude, y = latitude) +
-      ggraph::geom_edge_link(aes(width = weight), edge_colour = "grey") +
-      ggraph::geom_node_point(aes(size = eigenvector), colour = "purple") +
-      ggraph::geom_node_text(aes(label = LanName), repel = TRUE, size = 3) +
-      ggplot2::theme_void() +
-      ggplot2::labs(
-        title = paste("Graph for Type:", type_name),
-        size = "Eigenvector",
-        width = "Edge Weight"
-      ) +
-      ggplot2::scale_size_continuous(range = c(2, 10)) +
-      ggraph::scale_edge_width_continuous(range = c(0.5, 3)) +
-      ggraph::theme_graph(base_size = 12)
-
-    print(plot)
-  })
-```
-
-Save a graph for each of the trade items
-```{r}
-# Create a list of tidygraph objects, one for each "type"
-graph_list <- edge_list_summary |>
-  dplyr::group_split(type) |>
-  purrr::map(\(edges_for_type) {
-    # Get the type name
-    type_name <- unique(edges_for_type$type)
-
-    # Create a filtered tidygraph object for this type
-    graph_for_type <- tidygraph::tbl_graph(
-      nodes = node_list,
-      edges = edges_for_type,
-      directed = FALSE
-    ) |>
-      tidygraph::filter(!tidygraph::node_is_isolated()) |> # Remove isolated nodes
-      tidygraph::mutate(betweenness = tidygraph::centrality_betweenness()) # Add betweenness centrality
-
-    # Return the graph, named with the type
-    graph_for_type
-  }) |>
-  purrr::set_names(unique(edge_list_summary$type))
-
-# Save the list of graphs for further processing
-save(graph_list, file = "graph_list.RData")
-
+  plot_comm_graph(
+    graph,
+    community = "louvain",
+    title = item,
+    subtitle = "Louvain network communities"
+  )
+}
 ```
@@
 
+graph |> 
+  activate(nodes) |> 
+  filter(!LanName %in% c("Wolof", "Sereer", "Mandjak")) |> 
+plot_edgecost_graph("inv_cost", 20, "gold")
 ```
 
-```{r}
-#knock out WOlof from the gold network
-graph |> 
-  activate(edges) |> 
-  filter(type == "gold") |> 
-  activate(nodes) |> 
-  filter(LanName != "Wolof") |>
-  filter(!node_is_isolated()) |> 
-  plot_graph()
-```
+You're dealing with a weighted, undirected trade network where edge weights likely reflect trade volume or strength of connection, and node coordinates give geographic context.
+To find the most important routes, you can approach this problem from multiple complementary angles, depending on your goal.
 
-The graphs need to be sparsified now, as they are relatively useless as they are.
-```{r}
-library(backbone)
+Here are several analytical strategies, ranging from classical network analysis to spatially informed metrics:
+🧭 1. Weighted Shortest Paths (Trade Efficiency)
 
-disp <- disparity(graph_list$gold, alpha = 0.5)
-plot_graph(disp)
+    Goal: Find the most efficient trade routes considering both weight (stronger trade) and distance (shorter path).
 
-```
+    Method:
 
-```{r}
-# Find glott_code values in edge lists that are not in the node_list
-find_missing_glott_codes <- function(edgelists, node_list) {
-  # Combine all unique glott_code values from the edge lists
-  all_glott_codes <- lapply(edgelists, function(edges) {
-    unique(c(edges$glott_code_1, edges$glott_code_2))
-  }) |>
-    unlist() |>  # Flatten into a single vector
-    unique()     # Remove duplicates
+        Create a cost function that combines weight and geographic distance. For example:
 
-  # Find glott_codes not present in the node_list
-  missing_glott_codes <- setdiff(all_glott_codes, node_list$glott_code)
+        edge_cost = distance / weight
 
-  return(missing_glott_codes)
-}
+        Then use igraph::shortest_paths() or tidygraph::distances() to find paths that minimize this cost.
 
-# Apply the function
-missing_glott_codes <- find_missing_glott_codes(edgelists, node_list)
+    Geographic distance can be calculated using the Haversine formula or geosphere::distHaversine().
 
-# Display the missing glott_code values
-print(missing_glott_codes)
+📌 2. Edge Betweenness Centrality
 
-```
+    Goal: Find edges that are most frequently used in shortest paths — "key trade routes".
+
+    Method:
+
+    tidygraph::activate(edges) %>%
+      tidygraph::mutate(edge_betweenness = centrality_edge_betweenness(weights = cost))
+
+    You’ll need a cost column (e.g., inverse of weight or distance/weight) as edge weights for the centrality calculation.
+
+📈 3. Flow-Based Importance (Max Flow / Min Cut)
+
+    Goal: Assess trade capacity or bottlenecks.
+
+    Method:
+
+        Use igraph::max_flow() between key trade hubs (e.g., high-degree or central nodes).
+
+        This identifies routes with the greatest potential trade capacity.
+
+📊 4. Geographically Constrained MST (Minimum Spanning Tree)
+
+    Goal: Extract a simplified network showing the most critical, efficient paths covering the entire network.
+
+    Method:
+
+        Compute edge distances and weight them with inverse trade strength.
+
+        Use igraph::mst() on the resulting cost-weighted graph.
+
+    This highlights low-cost, high-value trade connections.
+
+🌐 5. Community Bridge Edges
+
+    Goal: Find routes that link otherwise separated regions.
+
+    Method:
+
+        Run community detection (e.g., Louvain).
+
+        Identify edges connecting nodes of different communities.
+
+        These are often interregional trade routes.
+
+🛰️ 6. Geographic Visualization & Filtering
+
+    Goal: Visually identify trade corridors.
+
+    Method:
+
+        Overlay your graph on a map using ggraph + ggplot2 with geographic coordinates.
+
+        Highlight high-weight, long-distance edges or edges with high betweenness.
 

### 2025-06-04 analysis/paper/loan word maps.qmd
static maps

@@
 ## Retrieve and clean data
+
 In order to have location data, we read in the African languages, their glottoglog code, and their location from the glottolog database (@hammarstromGlottolog2024).
@@
 ### Add further layers to the map
+
 Get the archaeological data layers
+
 ```{r}
@@
 function to add further layers
+
 ```{r}
@@
 
+## Make static maps
+
+```{r}
+library(rnaturalearth)
+
+# Load coastline and river data
+coastline <- ne_coastline(scale = "medium", returnclass = "sf")
+rivers <- ne_download(scale = "medium", type = "rivers_lake_centerlines", category = "physical", returnclass = "sf")
+```
+
+```{r}
+plot_language_map <- function(data, coastline, rivers, bbox_padding = 0.1) {
+  # Ensure consistent CRS across all inputs
+  common_crs <- sf::st_crs(coastline)  # Use the CRS of the coastline as reference
+  data <- sf::st_transform(data, crs = common_crs)
+  rivers <- sf::st_transform(rivers, crs = common_crs)
+
+  # Compute the bounding box of the data points
+  bbox <- sf::st_bbox(data)
+  # Extract numeric values for xmin, xmax, ymin, ymax to ensure clean handling
+  bbox_xmin <- as.numeric(bbox["xmin"])
+  bbox_xmax <- as.numeric(bbox["xmax"])
+  bbox_ymin <- as.numeric(bbox["ymin"])
+  bbox_ymax <- as.numeric(bbox["ymax"])
+
+  # Compute expanded bounding box
+  expanded_bbox <- sf::st_bbox(
+    c(
+      xmin = bbox_xmin - bbox_padding * (bbox_xmax - bbox_xmin),
+      xmax = bbox_xmax + bbox_padding * (bbox_xmax - bbox_xmin),
+      ymin = bbox_ymin - bbox_padding * (bbox_ymax - bbox_ymin),
+      ymax = bbox_ymax + bbox_padding * (bbox_ymax - bbox_ymin)
+    ),
+    crs = sf::st_crs(data)  # Use CRS from input data
+  )
+  # Convert bbox to polygon as sf object
+  expanded_bbox_sf <- sf::st_as_sfc(expanded_bbox)
+
+  # Crop coastline and rivers to match the bounding box
+  coastline_cropped <- sf::st_intersection(coastline, expanded_bbox_sf)
+  rivers_cropped <- sf::st_intersection(rivers, expanded_bbox_sf)
+
+  # Extract coordinates and drop geometry
+  data <- data |>
+    dplyr::mutate(
+      lon = sf::st_coordinates(geometry)[, 1],
+      lat = sf::st_coordinates(geometry)[, 2]
+    ) |>
+    sf::st_drop_geometry()
+
+  # Count occurrences of each category and order
+  category_order <- data |>
+    dplyr::count(style, name = "freq") |>
+    dplyr::arrange(dplyr::desc(freq)) |>
+    dplyr::mutate(color = dplyr::case_when(
+      dplyr::row_number() <= 12 ~ rcartocolor::carto_pal(n = 12, name = "Bold")[dplyr::row_number()], 
+      TRUE ~ "#3b3b3b"
+    ))
+
+  # Merge color mapping back to the data
+  data <- dplyr::left_join(data, category_order, by = "style")
+
+  # Plot the map
+  ggplot2::ggplot() +
+    ggplot2::geom_sf(data = coastline_cropped, color = "black", linewidth = 0.5) +  # Cropped coastline
+    ggplot2::geom_sf(data = rivers_cropped, color = "gray", linewidth = 0.3) +      # Cropped rivers
+    ggplot2::geom_text(
+      data = data,
+      ggplot2::aes(x = lon, y = lat, label = form, color = style),
+      position = ggplot2::position_jitter(width = 1.5, height = 1.5), 
+      size = 3
+    ) +  # Labels with jitter
+    ggplot2::scale_color_manual(values = setNames(category_order$color, category_order$style)) +  # Custom palette with dark gray fallback
+    ggplot2::coord_sf(
+      xlim = c(expanded_bbox["xmin"], expanded_bbox["xmax"]),
+      ylim = c(expanded_bbox["ymin"], expanded_bbox["ymax"]),
+      expand = FALSE
+    ) +  # Map extent controlled by `expanded_bbox`
+    ggplot2::theme_void() +
+    ggplot2::theme(legend.position = "none")
+}
+```
+
+```{r}
+plots <- purrr::map(data_list, ~plot_language_map(.x, coastline, rivers, bbox_padding = 0.15))
+```
+
+```{r}
+plot_language_map(data_list$gold, coastline, rivers, bbox_padding = 0.05)
+```
+
 ## Networks
+
 Next, we will create networks from this data, in order to be able to analyse and visualise connections.The dataset of languages is the nodelist, the edges are made on the basis of shared forms.
@@
 now, we read this into tidygraph network objects. First we take the language list and convert it into a usable edge list, and then make both into a network.
+
 ```{r}
@@
 
-Improve the gold graph:
-remove Wolof. It has the same form as the eastern Group, but its ouro form is probably portuguese influenced, whereas the lower Niger is not.
-Retain only stronger ties.
+Improve the gold graph: remove Wolof. It has the same form as the eastern Group, but its ouro form is probably portuguese influenced, whereas the lower Niger is not. Retain only stronger ties.
 
@@
 ### Fonio
+
 ```{r }
@@
 
-This graph is somewhat perplexing. What does the connection between Bangime (wīī-bɛ̀) and Igo (ā-wúbī,volta-Congo langue in southern Togo) mean? This is a question that needs further answering. 
+This graph is somewhat perplexing.
+What does the connection between Bangime (wīī-bɛ̀) and Igo (ā-wúbī,volta-Congo langue in southern Togo) mean?
+This is a question that needs further answering.
 For now, we can remove this pair, and concentrate on the rest.
@@
 ### Create a cost variable
+
 This combines weight and geographic distance.
+
 ```{r}
@@
 ### Via edge betweenness
-Goal: Find edges that are most frequently used in shortest paths — "key trade routes"
+
+Goal: Find edges that are most frequently used in shortest paths --- "key trade routes"
+
 ```{r}
@@
 
-Here are several analytical strategies, ranging from classical network analysis to spatially informed metrics:
-🧭 1. Weighted Shortest Paths (Trade Efficiency)
+Here are several analytical strategies, ranging from classical network analysis to spatially informed metrics: 🧭 1. Weighted Shortest Paths (Trade Efficiency)
 
+```         
 Goal: Find the most efficient trade routes considering both weight (stronger trade) and distance (shorter path).
@@
 Geographic distance can be calculated using the Haversine formula or geosphere::distHaversine().
+```
 
@@
 
+```         
 Goal: Find edges that are most frequently used in shortest paths — "key trade routes".
@@
 You’ll need a cost column (e.g., inverse of weight or distance/weight) as edge weights for the centrality calculation.
+```
 
@@
 
+```         
 Goal: Assess trade capacity or bottlenecks.
@@
     This identifies routes with the greatest potential trade capacity.
+```
 
@@
 
+```         
 Goal: Extract a simplified network showing the most critical, efficient paths covering the entire network.
@@
 This highlights low-cost, high-value trade connections.
+```
 
@@
 
+```         
 Goal: Find routes that link otherwise separated regions.
@@
     These are often interregional trade routes.
+```
 
@@
 
+```         
 Goal: Visually identify trade corridors.
@@
     Highlight high-weight, long-distance edges or edges with high betweenness.
+```
 

### 2025-06-05 README.qmd
static maps

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Loanwords and Trade in West Africa: grains, gold, salt, and iron"
 ```
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
-
-Our pre-print is online here:
-
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Schreiber, Henning & Gestrich, Nikolas (2025). _`r Title`_. Journal of global archaeology <https://doi.org/xxx/xxx>
 
@@
 
-  - [:file\_folder: paper](/analysis/paper): Quarto source document
-    for manuscript. Includes code to reproduce the figures and tables
-    generated by the analysis. It also has a rendered version,
-    `paper.docx`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+  - [:file\_folder: paper](/analysis/paper): Quarto source document. 
+    Includes code to reproduce the figures and tables
+    generated by the analysis.
   - [:file\_folder: data](/analysis/data): Data used in the analysis.
@@
     illustrations
-  - [:file\_folder:
-    supplementary-materials](/analysis/supplementary-materials):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
 

### 2025-06-05 analysis/paper/loan word maps.qmd
static maps

@@
 
+We need to standardise the "form" column to be lowercase
+
+```{r}
+data_list <- purrr::map(data_list, ~ {
+  .x |> 
+    dplyr::mutate(dplyr::across(.cols = starts_with("form"), ~ tolower(.)))
+})
+```
+
 ## Make interactive maps
@@
 #| fig-cap: "fonio word forms"
-# create_leaflet_map(data_list$iron, background = "bw") #uncomment to run
+
+ # create_leaflet_map(data_list$iron, background = "colour") #uncomment line to run
 ```
@@
 #| eval: false
-# save_maps(data_list)  #uncomment to run
+
+ save_maps(data_list, background = "colour")  #uncomment to run
+
 ```
@@
 library(rnaturalearth)
+library(rnaturalearthdata)
 
